Black-swan events in animal populations

### 2016-11-10 text/ms_pnas.tex

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}ajor surprises happen more often than expected in
financial, social, and natural systems \cite{taleb2007, sornette2009, may2008}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \cite{taleb2007}, fatalities from the largest wars
dwarf those from all others  \cite{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
ecological systems, background rates of global extinction are punctuated by
mass extinction \cite{harnik2012}, viruses can mutate suddenly to infect new
hosts, and billions of animals can die at once in mass mortality
events \cite{fey2015}.

@@

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{ Illustration of population dynamic models that allow for heavy
tails. (A, B) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (C to E)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels C and D represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (F to H)
Estimates of \(\nu\) from models fit to the times series in panels C to
E. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels F to H
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the extent to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (B)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

% than black-swan population
% increases and are usually driven by external events such as weather and
% disease. Therefore, ignoring the presence of black-swan events, and the
% predominance of crashes over increases, may underestimate extinction risk.
% TODO: add numbers, add skew add projections add length and Sigma

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012} including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction.
Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.
We expect that it is extrinsic forces that lead to the
prevalence of skew-$t$ distributed process noise:
there is no limit to how fast a population can decline due to a catastrophe,
but population increases are constrained
by the maximum biological rate of increase
possible for a given set of demographic traits
(e.g. population birth rate, mortality, age at maturity).

There are a number of caveats when considering the generality of our results.
The Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Additionally, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors. Finally, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.

@@

% todo the strongest correlates of finding blocks one events was the timeseries length
% ...implications

@@

TODO add assumptions here: assuming that productivity and density dependence are not time varying, this could be an important development for understanding the dynamics of longer time series

We chose the $t$ distribution for a number of reasons.
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations is absent.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \cite{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes
that could give rise to extreme events in population dynamics,
such as extreme climate events
\cite{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses
would be important for diagnosing the underlying causes
of extreme events and to develop predictive models,
it would be impossible to tease out each of these
underlying causes from population trend data alone.
Instead we seek a simple flexible phenomenological model
that can be applied across the widest possible
range of population time series to provide a
comparative overview of the prevalence
of black-swan events.

@@

We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again.
% These thresholds are in
% excess of the minimums recommended by Ref.~\cite{gelman2006a} of
% \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
% confidence intervals.
In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence.
This model is equivalent to a random walk with drift
model on log abundance, and therefore does not assume
that the timeseries is stationary.
Third, observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}ajor surprises happen more often than expected in
financial, social, and natural systems
\cite{sornette2002, albeverio2006, taleb2007, may2008, sornette2009}.
Massive unpredictable market swings are responsible for the majority of
financial gains and losses \cite{taleb2007}, fatalities from the largest wars
dwarf those from all others \cite{newman2005}, and the frequency of the most
damaging earthquakes has exceeded past expectations \cite{sornette2009}. In
ecological systems,
billions of animals can die at once in mass mortality
events \cite{fey2015},
background rates of global extinction are punctuated by
mass extinction \cite{harnik2012},
and evolution is characterized by
long periods of stasis punctuated by
bursts of speciation \cite{gould1977}.

@@

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\textwidth]{../analysis/t-nu-eg2.pdf}
\caption{Illustration of population dynamic models that allow for heavy
tails. (A, B) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (C to E)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels C and D represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (F to H)
Estimates of \(\nu\) from models fit to the times series in panels C to
E. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels F to H
illustrates the approximate region of heavy tails.}
\label{fig:1}
\end{figure*}

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the extent to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
As \(\nu\) approaches infinity, the $t$
distribution approaches the normal distribution (Fig.~1).

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (A to D)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel A are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.
}
\label{fig:2}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (A) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in A indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time-series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (B)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012}, including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction.
Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.
The prevalence of population die-offs rather than unexpected increases
is likely due to biology:
there is no limit to how fast a population can decline due to a catastrophe,
but population increases are constrained
by the maximum biological rate of increase
for a given set of demographic traits
(e.g. population birth rate, mortality, age at maturity).

There are a number of caveats when considering the generality of our results.
First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Second, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors (Supporting Information).
Third, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
Finally, our models assume that
the fitted parameters (e.g. productivity and density dependence)
do not vary over time.
Allowing for time varying parameters
could be an important development for
understanding the dynamics of longer time series.

@@

@@

We chose the $t$ distribution for a number of reasons.
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if evidence of heavy-tailed process deviations is absent.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \cite{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes
that could give rise to extreme events in population dynamics,
such as extreme climate events
\cite{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses
would be important for diagnosing the underlying causes
of extreme events and to develop predictive models,
it would be impossible to tease out each of these
underlying causes from population trend data alone.
Instead we seek a simple flexible phenomenological model
that can be applied across the widest possible
range of population time series to provide a
comparative overview of the prevalence
of black-swan events.

@@

We fit our models with Stan 2.12.1 \cite{hoffman2014, stan-manual2015, carpenter2016}, and R
3.3.1 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again.
In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence.
This model is equivalent to a random walk with drift
model on log abundance, and therefore does not assume
that the timeseries is stationary.
Third, observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Fourth, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).
% Our models also assume that their parameters
% (e.g. parameters representing productivity and density dependence)
